Unifying quasars with clumpy wind models

### 2015-08-19 paper.tex
adding CK changes slowly

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by M15. Previously (LK02, H13, M15),
we adopted a modified Saha approach to calculate the ionization fractions
of simple-ions. We have now improved our code to explicitly solve the 
rate equations in between ions in non-LTE. 
Not only does this
dispense with a number of small assumptions made in the modified Saha approach, 
it is also more numerically stable, 
and in principle allows the direct addition of extra physical 
processes such as Auger ionization, which would have to be approximated 
if using the previous technique.

To avoid approximating the complex AGN SED as e.g. a dilute blackbody \citep[][M15]{ML93},
we model the SED in a cell using the technique described by H13. In this scheme,
the mean intensity in a series of bands is modeled as a normalised power law in 
frequency $\nu$
\begin{equation}
J_{\nu,j}=K_{pl}\nu^{\alpha_{pl}},
\end{equation}
for a band $j$, or an exponential 
\begin{equation}
J_{\nu,j}=K_{exp} e^{(-h\nu/k T_{exp})}.
\end{equation}
Here, $K_{exp}$, $K_{pl}$, $T_{exp}$ and $\alpha_{pl}$ are spectral fit parameters
deduced from the band-limited radiation field estimators.
The ionization rate out of ion $i$ can then be written as 
\begin{equation}
R_{i,i+1}(J)= 
\displaystyle{
n_i \left(C_{i} n_e + 
\sum_{band~j=0}^{n}~
{\int_{\nu_j}^{~\nu_j+1}{J_{\nu,j}\sigma_i(\nu)\nu^{-1}d\nu}}
\right),}
\end{equation}
where $\sigma_i$ is the photoionization cross-section and $C_{i}$
represents the collisional ionization coefficient.
The recombination rate {\em into} ion $i$ is simply given by 
\begin{equation}
R_{i+1,i}(T_e) = (\alpha^i_{RR} + \alpha^i_{DR} + \alpha^i_{CR}) n_{i+1} n_e
\end{equation}
Where each $\alpha^i$ here is the recombination rate coefficient into the ground state of ion $i$.
The subscripts denote radiative, dielectronic and 
collisional (three-body) recombination respectively.
We thus neglect recombination to and from excited states in the simple-ion calculation. 
For simple-ions, we use a dilute Boltzmann equation to calculate 
the population of level $k$ in ionic stage $j$,
\begin{equation}
\frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R).
\end{equation}
Here $z_j$ is the partition function of ionic stage $j$,
$T_R$ is the effective radiation temperature,
$E_k$ is the energy difference between level $k$ and the ground state,
and $g_k$ is the statistical weight of level $k$.

@@

The accretion disc in our model is geometrically thin, but optically thick
and thus adopt a standard multi-temperature blackbody
using a \cite{shakurasunyaev1973} temperature profile.

The emergent SED is thus determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are assumed opaque, meaning
that photons which strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~R_G$.
The X-ray source in our model is treated as an isotropic sphere at the ISCO,
which emits r-packets according to a power law in flux with index $\alpha_X$ such that
\begin{equation}
F_X (\nu) = K \nu^{\alpha_X}.
\end{equation}
The normalisation, $K_X$ of this power law is such that it 
produces the specified 2-10~keV luminosity, $L_X$.
In addition to the disc and X-ray source, 
the wind is able to reprocess radiation. However, new 
photon packets are not produced in the wind (as in LK02). 
Instead, this reprocessing is dealt with by enforcing strict
radiative equilibrium ({\em modulo} adiabatic cooling)
via an indivisible energy packet
constraint (see Lucy 2002, M15; also section 2.3).
% \begin{equation}

@@

In this prescription, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$ at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind which accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

To take account of clumping in our outflow we adopt a simple parameterization
used in stellar wind modelling, known as {\em microclumping} \citep{hamann1998}. 
The key assumption here is that typical clump sizes
are much smaller than the typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. This approach is typically 
known as microclumping and allows one to introduce a `filling factor', $f$, which is the 
fraction of the volume of the plasma filled by clumps. We can then introduce the 
density enhancement, $D$, which is simply defined as 
\begin{equation}
D = \frac{1}{f}.
\end{equation}
We then multiply all densities in the model by $D$, and all emitting volumes
by $f$, meaning that all $\rho^2$ emissivities and opacities
(such as collisional excitation and recombination) will be enhanced, 
while all $\rho$ process (such as electron scattering and bound-free opacity)
remain unchanged for a given ionization state. 
% A factor $f$ is also
% applied to the opacities such that opacities which scale only with $\rho$ are not
% increased.

@@

Here we describe the results from our model, the parameters
of which are shown in table~1.
Parameters differing from the benchmark model of
H13 are highlighted with an asterisk.
This set of parameters
was arrived at by conducting a limited grid search over a 
5-dimensional parameter space involving the variables
$r_{min}$, $\theta_{min}$, $f$, $\alpha$ and $R_V$.
The full grid, including output spectral files and plots can be found at
\url{jhmatthews.github.io/quasar-wind-grid/}.
We evaluated these models qualitavely based on the following
criteria:
\begin{itemize}
\item Does the model maintain the correct ionization state and produce strong BALs?
\item Does strong line emission emerge at low inclinations?
\item Do H recombination lines appear in the spectrum?
\item Do a certain range of angles produce LoBAL features?
\item Does the model compare favourably to quasar composite spectra?
\end{itemize}
We arrived at a model which is one of the most promising,
but also broadly representative of a family of models. 
Examination of this individual model allows us to gain insight
into fundamental geometrical and physical constraints, 
as discussed in section 5, and assess the models potential
for unification.

@@

Figure~\ref{fig:wind} shows the physical properties of the wind.
The wind rises slowly from the disc at first, with clumped densities
of $n_H \sim 10^{11}~\rm{cm^{-3}}$ close to the disc plane.
The flow then accelerates over a scale length of $R_V=10^{19}~\rm{cm}$
up to a terminal velocity of around $3$ times the escape velocity 
($\sim10,000~\rm{km~s^{-1}}$). This gradual acceleration means that
the wind exhibits a stratified ionization structure, with low ionization material
in the base of the wind giving way to highly ionized plasma further out.
By clumping the wind, we are able to produce the range of ionization states observed
in wuasars and BALQSOs, while adopting a realistic $2-10$ keV X-ray luminosity
of $L_{X}=10^{45}~\rm{ergs~s^{-1}}$. Without clumping, this wind would be over-ionized 
to the extent that opacities in e.g., \civ\ would be entirely negligible (see H13).

One commonly used measure of the ionization state is the ionization parameter, $U$, given by
\begin{equation}
U = \frac{4\pi}{n_H c}\int_{13.6{\rm{eV}}}^{\infty}\frac{{J_{\nu}d\nu}}{h\nu}.
\end{equation}
\noindent where $\rm{n_H}$ is the local number density of H, and $\nu$ denotes photon 
frequency. The ionization parameter represents the ratio of the number density of 
ionizing photons to the local H density. It is however, a poor measure of the 
ionization state of the resonance species such as \civ\ as it encodes no information
about the shape of the SED. In this case the X-ray photons 
are dominant in the photoionization of the UV resonance line ions. 
This explains why a factor of 100 increase in X-ray luminosity requires
a clumping factor of 0.01, even though the value of U decreases by only a factor of $\sim10$ 
compared to H13. This is also the reason for significant Lyman edge photoabsorption
at the highest inclinations (see section 4.3).

Clumping also causes the total line luminosity to increase dramatically,
as recombination and collisional excitation are both proportional to
$n_e^2$. This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the X-rays heating the plasma.
The radii of these emitting regions is important,
and can be compared to observations. As shown in ~\ref{fig:wind},
the \civ\ line in our model is typically formed between 
$100-1000~R_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civ\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

The BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At lower inclinations, the strongest absorption occurs at the red edge,
whereas at high inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008).
In addition, the line profile shape is strongly dependent 
on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the shape of the line.
% the line profile shape. Figure ?? shows selection of line profiles 
% for different kinematic parameters to demonstrate this. 
% The degeneracy between parameters, and lack of available
% physical and observational constraints goes someway to showing why individual 
% fitting of line profile shapes with kinematic models is such a difficult
% exercise.

We find that the model can produce significant line emission
at low inclinations, particular in \civ, 
and the improved treatment of recombination results in a strong \la\
line. In the context of unification, this is a promising results, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles.
However, the red wing of the BAL profiles 
is generally stronger than seen in BALQSO spectra and composites.
This is due to a combination of strong collisionally excited emission
and continuum attenuation. To illustrate the amount of 
continuum attenuation we also show the continuum from a disc-only model.
The line-to-continuum ratios at low inclinations are also
significantly affected by disc foreshortening and limb darkening.
The angular distribution of the disc radiation is clearly
crucially important in determining the emergent line equivalent widths 
(EWs) -- this is discussed further in section~5.

To assess the ability of the model to match real 
quasar spectra, we also show {\sl Sloan Digital Sky Survey} (SDSS) quasar composites from
\cite{reichard2003}, normalised to the flux at 2000\AA\ in each panel.
The model matches the \civ\ and \mg\ line profiles fairly well, 
but the \la\ profile is a factor of $\sim2$ weaker than in the composite spectrum.

This is implies that a larger recombination line formation region is required in order 
to successfully reproduce the strong \la\ emission seen in 
all quasars.

>>>>

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by M15. Previously (LK02, H13, M15),
we adopted a modified Saha approach to calculate the ionization fractions
of simple-ions. We have now improved our code to explicitly solve the 
rate equations between ions in non-LTE. 
This dispenses with a number of small assumptions made in the modified Saha approach, 
is more numerically stable, 
and, in principle, allows the direct addition of extra physical 
processes such as Auger ionization.

%To avoid approximating the complex AGN SED as e.g. a dilute blackbody \citep[][M15]{ML93},
We model the SED in a cell using the technique described by H13. In this scheme,
the mean intensity in a series of bands is modeled as a normalised power law in 
frequency $\nu$
\begin{equation}
J_{\nu,j}=K_{pl}\nu^{\alpha_{pl}},
\end{equation}
for a band $j$, or an exponential 
\begin{equation}
J_{\nu,j}=K_{exp} e^{(-h\nu/k T_{exp})}.
\end{equation}
Here, $K_{exp}$, $K_{pl}$, $T_{exp}$ and $\alpha_{pl}$ are spectral fit parameters
deduced from the band-limited radiation field estimators.
The ionization rate out of ion $i$ can then be written as 
\begin{equation}
R_{i,i+1}(J)= 
\displaystyle{
n_i \left(C_{i} n_e + 
\sum_{band~j=0}^{n}~
{\int_{\nu_j}^{~\nu_j+1}{J_{\nu,j}\sigma_i(\nu)\nu^{-1}d\nu}}
\right),}
\end{equation}
where $\sigma_i$ is the photoionization cross-section and $C_{i}$
represents the collisional ionization coefficient.
The recombination rate {\em into} ion $i$ is given by 
\begin{equation}
R_{i+1,i}(T_e) = (\alpha^i_{RR} + \alpha^i_{DR} + \alpha^i_{CR}) n_{i+1} n_e,
\end{equation}
where each $\alpha^i$ here is the recombination rate coefficient into the ground state of ion $i$.
The subscripts denote radiative, dielectronic and 
collisional (three-body) recombination, respectively.
We neglect recombination to and from excited states in the simple-ion calculation. 
For simple-ions, we use a dilute Boltzmann equation to calculate 
the population of level $k$ in ionic stage $j$,
\begin{equation}
\frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R).
\end{equation}
Here $z_j$ is the partition function of ionic stage $j$,
$T_R$ is the effective radiation temperature,
$E_k$ is the energy difference between level $k$ and the ground state,
and $g_k$ is the statistical weight of level $k$.

@@

The accretion disc in our model is geometrically thin, but optically thick.
We thus adopt a standard multi-temperature blackbody
using a \cite{shakurasunyaev1973} temperature profile.

The emergent SED is then determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are assumed to beopaque, meaning
that photons which strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~R_G$.
The X-ray source in our model is treated as an isotropic sphere at the ISCO,
which emits r-packets according to a power law in flux with index $\alpha_X$, of the form
\begin{equation}
F_X (\nu) = K \nu^{\alpha_X}.
\end{equation}
The normalisation, $K_X$ of this power law is such that it 
produces the specified 2-10~keV luminosity, $L_X$.
In addition to the disc and X-ray source, 
the wind is able to reprocess radiation. However, new 
photon packets are not produced in the wind (as in LK02). 
Instead, this reprocessing is dealt with by enforcing strict
radiative equilibrium ({\em modulo} adiabatic cooling)
via an indivisible energy packet
constraint (see Lucy 2002, M15; also section 2.3).
% \begin{equation}

@@

In our model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind which accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

To allow for clumping in our outflow we adopt a simple approximation
used extensively in stellar wind modelling, known as 
{\em microclumping} \citep{hamann1998}. The key assumption here is that 
typical clump sizes are much smaller than the typical photon mean free path, 
and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `filling factor', $f$, which is the 
fraction of the volume of the plasma filled by clumps. The intra-clump
medium is assumed to be a vacuum. We can then introduce the 
density enhancement, $D$, which is defined as 
\begin{equation}
D = \frac{1}{f}.
\end{equation}
We then multiply all densities in the model by $D$, and all emitting volumes
by $f$, meaning that all $\rho^2$ emissivities and opacities
(such as collisional excitation and recombination) will be enhanced, 
while all $\rho$ process (such as electron scattering and bound-free opacity)
remain unchanged for a given ionization state. 
% A factor $f$ is also
% applied to the opacities such that opacities which scale only with $\rho$ are not
% increased.

@@

Here we describe the results from our model, the parameters
of which are shown in table~1.
Parameters differing from the benchmark model of
H13 are highlighted with an asterisk.
This set of parameters
was arrived at by conducting a limited grid search over a 
5-dimensional parameter space involving the variables
$r_{min}$, $\theta_{min}$, $f$, $\alpha$ and $R_V$.
The full grid, including output spectral files and plots can be found at
\url{jhmatthews.github.io/quasar-wind-grid/}.
We evaluated these models qualitatively based on the following
criteria:
\begin{itemize}
\item Does the model maintain the correct ionization state and produce strong BALs?
\item Does strong line emission emerge at low inclinations?
\item Do H recombination lines appear in the spectrum?
\item Do a certain range of angles produce LoBAL features?
\item Does the model compare favourably to quasar composite spectra?
\end{itemize}
We arrived at a model which is one of the most promising,
but also broadly representative of a family of models. 
Examination of this individual model allows us to gain insight
into fundamental geometrical and physical constraints, 
as discussed in section 5, and assess the models potential
for unification.

@@

Figure~\ref{fig:wind} shows the physical properties of the wind.
The wind rises slowly from the disc at first, with clumped densities
of $n_H \sim 10^{11}~\rm{cm^{-3}}$ close to the disc plane.
The flow then accelerates over a scale length of $R_V=10^{19}~\rm{cm}$
up to a terminal velocity equal to the escape velocity at the streamline base
($\sim10,000~\rm{km~s^{-1}}$). This gradual acceleration means that
the wind exhibits a stratified ionization structure, with low ionization material
in the base of the wind giving way to highly ionized plasma further out.
By clumping the wind, we are able to produce the range of ionization states observed
in quasars and BALQSOs, while adopting a realistic $2-10$ keV X-ray luminosity
of $L_{X}=10^{45}~\rm{ergs~s^{-1}}$. Without clumping, this wind would be over-ionized 
to the extent that opacities in e.g., \civ\ would be entirely negligible (see H13).

One commonly used measure of the ionization state is the ionization parameter, $U_H$, given by
\begin{equation}
U_H = \frac{4\pi}{n_H c}\int_{13.6{\rm{eV}}}^{\infty}\frac{{J_{\nu}d\nu}}{h\nu}.
\end{equation}
\noindent where $\rm{n_H}$ is the local number density of H, and $\nu$ denotes photon 
frequency. The ionization parameter represents the ratio of the number density of 
H ionizing photons to the local H density. It is, however, a poor measure of the 
ionization state of species such as \civ\ as it encodes no information
about the shape of the SED. In our case, the X-ray photons 
are dominant in the photoionization of the UV resonance line ions. 
This explains why a factor of 100 increase in X-ray luminosity requires
a clumping factor of 0.01, even though the value of $U_H$ decreases by only a factor of $\sim10$ 
compared to H13. This is also the reason for significant Lyman edge photoabsorption
at the highest inclinations (see section 4.3).

Clumping also causes the total line luminosity to increase dramatically,
as recombination and collisional excitation are both proportional to
$n_e^2$. This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the X-rays heating the plasma.
The radii of these emitting regions is important,
and can be compared to observations. As shown in figure~\ref{fig:wind},
the \civ\ line in our model is typically formed between 
$100-1000~R_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civ\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At lower inclinations, the strongest absorption occurs at the red edge,
whereas at high inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008).
In addition, the line profile shape is strongly dependent 
on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the shape of the line.
% the line profile shape. Figure ?? shows selection of line profiles 
% for different kinematic parameters to demonstrate this. 
% The degeneracy between parameters, and lack of available
% physical and observational constraints goes someway to showing why individual 
% fitting of line profile shapes with kinematic models is such a difficult
% exercise.

We find that the model can produce significant line emission
at low inclinations, particular in \civ, 
and the improved treatment of recombination results in a strong \la\
line. In the context of unification, this is a promising result, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles.
However, the red wing of the BAL profiles 
is generally stronger than seen in BALQSO spectra and composites.
This is due to a combination of strong collisionally excited emission
and continuum attenuation. To illustrate the amount of 
continuum attenuation we also show the continuum from a disc-only model.
The line-to-continuum ratios at low inclinations are also
significantly affected by disc foreshortening and limb darkening.
The angular distribution of the disc radiation is clearly
crucially important in determining the emergent line equivalent widths 
(EWs) -- this is discussed further in section~5.

To assess the ability of the model to match real 
quasar spectra, we also show {\sl Sloan Digital Sky Survey} (SDSS) quasar composites from
\cite{reichard2003}, normalised to the flux at 2000\AA\ in each panel.
The model matches the \civ\ and \mg\ line profiles fairly well, 
but the \la\ profile is a factor of $\sim2$ weaker than in the composite spectrum.
This is one of the main weaknesses of our models.
